---
title: "Exploratory Analyses: Overview"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(lubridate)
library(scales)
library(viridis)

mvc_clean <- read_csv("E:/Desktop/mvc_clean.csv", show_col_types = FALSE)
```

# Exploratory Analyses: Overview

This section provides a descriptive overview of NYC motor vehicle collisions (MVCs), including borough patterns, severity distribution, and temporal trends.

---

# 1. Collision Counts by Borough

```{r}
mvc_clean |>
  count(borough, name = "n") |>
  ggplot(aes(x = reorder(borough, n), y = n, fill = borough)) +
  geom_col() +
  scale_fill_viridis_d() +
  coord_flip() +
  labs(
    title = "Total Collisions by Borough",
    x = "Borough",
    y = "Number of Collisions"
  ) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none")
```

---

# 2. Severity Distribution

```{r}
mvc_clean |>
  mutate(
    severity = case_when(
      number_of_persons_killed > 0 ~ "Fatal",
      number_of_persons_injured > 0 ~ "Injury",
      TRUE ~ "Property Damage Only"
    )
  ) |>
  count(severity) |>
  ggplot(aes(x = severity, y = n, fill = severity)) +
  geom_col() +
  scale_fill_viridis_d() +
  labs(
    title = "Collision Severity Distribution",
    x = NULL,
    y = "Number of Crashes"
  ) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none")
```

---

# 3. Collisions by Hour of Day

```{r}
mvc_clean |>
  mutate(hour = hour(crash_time)) |>
  count(hour) |>
  ggplot(aes(x = hour, y = n)) +
  geom_line(color = "#1f78b4", size = 1.2) +
  geom_point(color = "#084c7f", size = 2) +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Collisions by Hour of Day",
    x = "Hour",
    y = "Number of Crashes"
  ) +
  theme_minimal(base_size = 15)
```

---

# 4. Collisions by Day of Week

```{r}
mvc_clean |>
  mutate(weekday = wday(crash_date, label = TRUE)) |>
  count(weekday) |>
  ggplot(aes(x = weekday, y = n, fill = weekday)) +
  geom_col() +
  scale_fill_viridis_d() +
  labs(
    title = "Collisions by Day of Week",
    x = NULL,
    y = "Number of Crashes"
  ) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none")
```

---

# 5. Monthly Trend Over Time

```{r}
mvc_clean |>
  count(crash_month) |>
  ggplot(aes(x = crash_month, y = n)) +
  geom_line(color = "#2C93E8", size = 1.1) +
  geom_smooth(se = FALSE, color = "#084c7f") +
  labs(
    title = "Monthly Collision Trend Over Time",
    x = "Month",
    y = "Number of Collisions"
  ) +
  theme_minimal(base_size = 15)
```

---

# 6. Heatmap: Hour × Weekday

```{r}
mvc_clean |>
  mutate(
    hour = hour(crash_time),
    weekday = wday(crash_date, label = TRUE)
  ) |>
  count(weekday, hour) |>
  ggplot(aes(x = hour, y = weekday, fill = n)) +
  geom_tile() +
  scale_fill_viridis(option = "C") +
  labs(
    title = "Heatmap of Collisions by Hour and Day of Week",
    x = "Hour of Day",
    y = "Day of Week",
    fill = "Collisions"
  ) +
  theme_minimal(base_size = 15)
```

---

# 7. Top 10 Contributing Factors

```{r}
mvc_clean |>
  count(contributing_factor_vehicle_1) |>
  arrange(desc(n)) |>
  slice_head(n = 10) |>
  ggplot(aes(x = reorder(contributing_factor_vehicle_1, n), y = n)) +
  geom_col(fill = "#1f78b4") +
  coord_flip() +
  labs(
    title = "Top 10 Contributing Factors",
    x = "Contributing Factor",
    y = "Count"
  ) +
  theme_minimal(base_size = 15)
```

---

# 8. Top 10 Vehicle Types in Crashes

```{r}
mvc_clean |>
  count(vehicle_type_code_1) |>
  arrange(desc(n)) |>
  slice_head(n = 10) |>
  ggplot(aes(x = reorder(vehicle_type_code_1, n), y = n)) +
  geom_col(fill = "#33a02c") +
  coord_flip() +
  labs(
    title = "Top 10 Vehicle Types in Crashes",
    x = "Vehicle Type",
    y = "Count"
  ) +
  theme_minimal(base_size = 15)
```

---

# 9. Hourly Injuries vs Fatalities

```{r}
mvc_clean |>
  mutate(
    hour = hour(crash_time),
    severity = case_when(
      number_of_persons_killed > 0 ~ "Fatal",
      number_of_persons_injured > 0 ~ "Injury",
      TRUE ~ "PDO"
    )
  ) |>
  count(hour, severity) |>
  ggplot(aes(x = hour, y = n, color = severity)) +
  geom_line(size = 1.1) +
  scale_color_viridis_d(direction = -1) +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Hourly Pattern of Injuries and Fatalities",
    x = "Hour",
    y = "Count",
    color = "Severity"
  ) +
  theme_minimal(base_size = 15)
```

---

# Summary

These exploratory analyses reveal consistent spatial differences across boroughs, strong temporal patterns (especially during rush hours and weekdays), and clear distributions of injury severity. These results guide more advanced analyses in subsequent sections.
