---
title: "Exploratory Analyses: Time Trends"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)
library(lubridate)
library(scales)
library(viridis)
mvc_clean <- read_csv("data/mvc_clean.csv", show_col_types = FALSE)
```

# Time Trends

This section examines temporal patterns of motor vehicle collisions in New York City across different time scales: hourly, daily, monthly, and yearly.

---

# Monthly Trends

```{r monthly}
mvc_clean <- mvc_raw |>
  clean_names() |>
  mutate(
    crash_date = as.character(crash_date),
    crash_time = as.character(crash_time),

    crash_datetime = suppressWarnings(
      ymd_hm(paste(crash_date, crash_time))
    ),

    year  = year(crash_datetime),
    month = month(crash_datetime, label = TRUE),
    wday  = wday(crash_datetime, label = TRUE),
    hour  = hour(crash_datetime)
  ) |>
  filter(
    !is.na(crash_datetime),      # 🚀 必须
    !is.na(latitude),
    !is.na(longitude),
    !is.na(borough)
  )

```

**Interpretation:**  
Monthly counts often show seasonal patterns, with variations that may correspond to weather, traffic volume, or policy changes.

---

# Yearly Trends

```{r yearly}
mvc_clean |>
  count(year) |>
  ggplot(aes(x = year, y = n)) +
  geom_line(size = 1.2, color = "#1a5ca0") +
  geom_point(size = 2, color = "#2C93E8") +
  scale_x_continuous(breaks = unique(mvc_clean$year)) +
  labs(
    title = "Annual Collision Totals",
    x = "Year",
    y = "Number of Crashes"
  ) +
  theme_minimal(base_size = 14)
```

**Interpretation:**  
Long-term variation may reflect population changes, infrastructure alteration, or shocks such as COVID-era mobility reductions.

---

# Day of Week

```{r weekday}
mvc_clean |>
  count(wday) |>
  ggplot(aes(x = wday, y = n)) +
  geom_col(fill = "#2C93E8") +
  labs(
    title = "Collisions by Day of Week",
    x = "Day of Week",
    y = "Number of Crashes"
  ) +
  theme_minimal(base_size = 14)
```

**Interpretation:**  
Weekday traffic patterns produce higher collision counts Monday–Friday, while weekends show lower totals.

---

# Hour of Day

```{r hour}
mvc_clean |>
  mutate(hour = as.integer(hour)) |>
  filter(!is.na(hour), hour >= 0, hour <= 23) |>
  count(hour) |>
  ggplot(aes(x = hour, y = n)) +
  geom_line(color = "#2C93E8", size = 1.2) +
  geom_point(size = 2, color = "#1a5ca0") +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Collisions by Hour of Day",
    x = "Hour",
    y = "Number of Crashes"
  ) +
  theme_minimal(base_size = 14)
```

**Interpretation:**  
Collision activity spikes during commuting hours (8–10 AM and 4–6 PM).

---

# Hour by Day of Week Heatmap (Advanced)

This heatmap visualizes joint patterns across **hour of day** and **day of week**.

```{r heatmap}
mvc_clean |>
  mutate(
    hour = as.integer(hour),
    wday = factor(wday, levels = c("Sun","Mon","Tue","Wed","Thu","Fri","Sat"))
  ) |>
  filter(!is.na(hour), hour >= 0, hour <= 23) |>
  count(wday, hour) |>
  ggplot(aes(x = hour, y = wday, fill = n)) +
  geom_tile() +
  scale_fill_viridis(option = "C", direction = -1) +
  labs(
    title = "Collisions by Hour and Day of Week",
    x = "Hour of Day",
    y = "Day of Week",
    fill = "Count"
  ) +
  theme_minimal(base_size = 14)
```

**Interpretation:**  
The heatmap reveals:

- Morning and evening rush peaks across all weekdays  
- Lower overnight crash counts  
- Saturday afternoon shows elevated activity compared to Sunday  

---

# Summary

These analyses reveal:

- Strong diurnal (hourly) patterns  
- Weekday vs weekend differences  
- Monthly and annual variation  
- Clear joint structure in hour × weekday patterns  

These insights guide later modeling and help contextualize borough-level and spatial analyses.
