---
title: "Mapping Collisions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(leaflet)
library(leaflet.extras)
library(viridis)
library(htmltools)

mvc_clean <- read_csv("E:/Desktop/mvc_clean.csv", show_col_types = FALSE)
```

# Mapping Motor Vehicle Collisions in NYC

This section visualizes the geographic distribution of collisions across New York City.

---

# ⚠ Data Note

We use latitude and longitude directly from the cleaned dataset.

---

# 1. Basic Leaflet Map (5,000 samples)

```{r leaflet-sample}
set.seed(123)

mvc_sample <- mvc_clean |>
  filter(!is.na(latitude), !is.na(longitude)) |>
  slice_sample(n = 5000)

leaflet(mvc_sample) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  setView(lng = -73.97, lat = 40.75, zoom = 11) |>
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.5,
    color = "#2C93E8"
  )
```

---

# 2. Severity Map (Property Damage / Injury / Fatal)

```{r severity-map}
mvc_sample2 <- mvc_clean |>
  mutate(
    severity = case_when(
      number_of_persons_killed > 0 ~ "Fatal",
      number_of_persons_injured > 0 ~ "Injury",
      TRUE ~ "Property Damage Only"
    )
  ) |>
  slice_sample(n = 5000)

pal <- colorFactor(
  palette = c("#7db7ff", "#2C93E8", "#FF5733"),
  domain = c("Property Damage Only", "Injury", "Fatal")
)

leaflet(mvc_sample2) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  setView(lng = -73.97, lat = 40.75, zoom = 11) |>
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.6,
    color = ~pal(severity),
    label = ~htmlEscape(severity)
  ) |>
  addLegend(
    position = "bottomright",
    pal = pal,
    values = mvc_sample2$severity,
    title = "Severity"
  )
```

---

# 3. Heat Map (Collision Density)

```{r heatmap}
leaflet(mvc_sample) |>
  addProviderTiles(providers$CartoDB.DarkMatter) |>
  setView(lng = -73.97, lat = 40.75, zoom = 11) |>
  addHeatmap(
    lng = ~longitude,
    lat = ~latitude,
    blur = 20,
    max = 0.05,
    radius = 10
  )
```

---

# 4. Borough Faceting (Static Plot)

```{r faceted}
mvc_clean |>
  filter(!is.na(borough)) |>
  ggplot(aes(x = longitude, y = latitude, color = borough)) +
  geom_point(alpha = 0.2, size = 0.3) +
  facet_wrap(~ borough) +
  scale_color_viridis_d() +
  coord_equal() +
  theme_minimal(base_size = 13) +
  labs(
    title = "Collision Locations by Borough",
    x = "Longitude",
    y = "Latitude"
  )
```

---

# Summary

Spatial patterns reveal:

- Hotspots in Manhattan, Brooklyn, Western Queens  
- Severity clustering along major roadways  
- Borough-level differences in spatial spread  

Mapping provides vital context for subsequent statistical modeling.
